--- 
title: "Notes on Modern Statistics for the Social and Behavioral Sciences (MSSBS)"
author: "Peter Baumgartner"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::html_document:
    df_print: paged
  pdf_document: default
description: 'This book accompanies "Rand, W. (2017). Modern Statistics for the Social and Behavioral Sciences: A Practical Introduction, Second Edition (New edition). Boca Raton, FL: Taylor & Francis Inc.". My intention is to learn and explore the concepts and procedures taught from this introductory book. Please keep in mind that this book is just a kind of a training exercise for me. There will be no new insights presented and as I am still learning the basics of R and statistics you may find minsunderstandings and errors in my writings. For authoritative reference you have the consult the original book.'
documentclass: book
bibliography:
- book.bib
- packages.bib
biblio-style: apalike
link-citations: yes
lof: True
lot: True
---

# Introduction


```{r setup, echo=F}

### setting up working environment
### for details see: https://yihui.name/knitr/options/
knitr::opts_chunk$set(
        echo = T,
        message = T,
        error = T,
        warning = T,
        comment = '##',
        highlight = T,
        prompt = T,
        strip.white = T,
        tidy = T
        )

###########################################################
cor_list = list()
def_list = list()
exr_list = list()
exm_list = list()
lem_list = list()
thm_list = list()
knitr::knit_hooks$set(engine = function(before, options) {
    if ( before ) {
        if ( options$engine == "theorem" ) {
            thm_list[[options$label]] <<- options$name
        } else if ( options$engine == "definition" ) {
            def_list[[options$label]] <<- options$name
        } else if ( options$engine == "corollary" ) {
            cor_list[[options$label]] <<- options$name
        } else if ( options$engine == "exercise" ) {
            exr_list[[options$label]] <<- options$name
        } else if ( options$engine == "lemma" ) {
            lem_list[[options$label]] <<- options$name
        } else if ( options$engine == "example" ) {
            exm_list[[options$label]] <<- options$name
        }
    }
    NULL
}) 

#######################################################################
# load the huge R file with the special wilcox functions
# prepare separate environment to prevent clogging of workspace
allfun <- new.env()
with(allfun, {source(file = "_mssbs-software/Rallfun.R", local = TRUE)})


#######################################################################
## Miscellanous other general settings

# disabling scientific notation in calculation
options("scipen" = 999, "digits" = 1) 

```

This chapter presents my motiviation to write this book besides some requirements and starting points so that you can follow the following chapters on your own.

## Status of this book

This book collects my excerpts and personal notes on 

> Rand, W. (2017). Modern Statistics for the Social and Behavioral Sciences: A Practical Introduction, Second Edition (2 New edition). Boca Raton, FL: Taylor & Francis Inc.

Please keep in mind that this book is just a kind of a training exercise for me. There will be no new insights presented and as I am still learning the basics of R and statistics you may find minsunderstandings and errors in my writings. For authoritative reference you have to consult the original book.  

With this book I pursue three personal learning goals:

* To get more practice with R, the free software environment for statistical computing and graphics.
* To learn more about modern statistics for the social sciences to get at least an intermediate practical knowledge for developing my own data analysis projects.
* To get more experience with the bookdown package.

## Web pages

The book (and therefore my notes as well) uses the following resources:

* **Homepage:** https://dornsife.usc.edu/cf/labs/wilcox/wilcox-faculty-display.cfm
* **Data:** https://dornsife.usc.edu/labs/rwilcox/datasets/
* **Software:** https://dornsife.usc.edu/labs/rwilcox/software/
* **Solutions:** https://dornsife.usc.edu/assets/sites/239/docs/CRC_answers.pdf
* **Errata**: https://dornsife.usc.edu/assets/sites/239/docs/CRC_2nd_Ed_errata.pdf
* **GitHub WRS-Text**: https://dornsife.usc.edu/assets/sites/239/docs/Rallfun-v37.txt (Sept. 2019)
* **GitHub WRS-Github**: https://github.com/nicebread/WRS/tree/master/pkg (v0.36, June 2016)
* **GitHub WRS2-CRAN**: https://github.com/cran/WRS2/tree/master/R (v0.36, 2019-06-05, only selection of functions but with Help Pages)
* **GitHub WRS3**: https://github.com/mrxiaohe/WRScpp (couldn't install: error: cannot open file 'man': No such file or directory)
* **New Functions**: https://dornsife.usc.edu/assets/sites/239/docs/updatev4_info.pdf

## Software (R Packages)

To use the software accompanying the book is a little bit complicated as there exist three different options each with their advantages and disadvantages.

### Textfile with all functions

The current release (Sept. 2019) `Rallfun-v37.txt` is a [textfile](https://dornsife.usc.edu/assets/sites/239/docs/Rallfun-v37.txt) and should be sourced with `source_url("https://dornsife.usc.edu/assets/sites/239/docs/Rallfun-v37.txt")`. It is the most updated version. It has the disadvantage that no help files are available. But most functions called without parenthesis from the console display the source code which includes on the top a summarized information about the function and its parameters..

I have converted the `.txt` extension to an `.R` file. Loading `Rallfun-v37.R` into RStudio displayed a message that several packages (with their dependencies) needed to be installed:

> install.packages(c("ddalpha", "gbm", "ICSOutlier", "mboost", "mrfDepth", "neuralnet", "rankFD", "tseries"))
also installing the dependencies ‘TTR’, ‘ICS’, ‘stabs’, ‘nnls’, ‘Deriv’, ‘quantmod’

To source the huge file will take some time and generates an error: `Error in eval(ei, envir) : object 'splotg5' not found`. Version v36 has still worked. It turned out that the function `splotg5()` in v37 was missing. After I copied it from version v36 the error message disappeared.

The many functions not only cluttered up the global environment (workspace) but needed too much of the working memory too.  Memory problems of this huge file (> 2 MB) practically prevents working with it. So I have loaded the file in the setup chunk in a local environment named `allfun` to prevent clogging the workspace (the Global Environment). Calling one of these special function provided by Wilcox I have to embed the R code into `with(allfun, {x})`, where `x` stands for the embedded R code.


### GitHub Version

You can download the `WRS Package` from https://github.com/nicebread/WRS from Github, where you can also find the details of the installation process. (I am not sure, but the installation process seems more cumbersome --  at least when you have to download all the dependent and suggested packages manually. But I believe most of these packages would be installed automatically.)

The Github version is not always up-to-date, but at the time of this writing (2019-08-03), it has already installed Version v0.36. On Github, you will also find instructions on the installation of the C++ version. The C++ functions in the package `WRScpp` may considerably shorten the executions times of some functions. But currently I have problems to install them. (See my [error report](https://github.com/mrxiaohe/WRScpp/issues/5).)


```{r install-packages, warning=F, message=F, cache=F, echo=F}

list.of.packages <- c("MASS", "akima", "robustbase", "devtools", "remote",
                  "cobs", "robust", "mgcv", "scatterplot3d", "quantreg", 
                  "rrcov", "lars", "pwr", "trimcluster", "mc2d",
                  "psych", "Rfit", "DepthProc", "class", "fda",
                  "RcppArmadillo", # above functions required by Wilcox' file
                  "WRS2", # Wilcox package on CRAN, necessary for its data files
                  "xfun", "knitr", "tidyverse", "kableExtra") # functions for my work on this book
packages.to.install <-
  list.of.packages[!(list.of.packages %in% installed.packages()[, "Package"])]
if (length(packages.to.install) > 0) install.packages(packages.to.install)
packages.to.update <-
  list.of.packages[(list.of.packages %in% old.packages()[, "Package"])]
if (length(packages.to.update) > 0) update.packages(oldPkgs = packages.to.update)

# install the WRS package from GitHub
# it contains some of the Wilcox functions `allfun` with help files
# loaded after WRS2 (from CRAN) because it is more up to date
# so that many duplicate functions from WRS2 will be masked
if (!requireNamespace("WRS"))
  remotes::install_github(
    "nicebread/WRS",
    subdir = "pkg",
    build_manual = TRUE,
    build_vignettes = TRUE
  )

# load all packages above
xfun::pkg_attach2(list.of.packages, "WRS")

#######################################################################
# Store all variables up to this line
# Later generated variables will be deleted after end of each chapter
freeze <- ls()


```



### Install CRAN version

The appeal of this limited edition of 37 functions are the included help files. There is an long article^[Mair, P., & Wilcox, R. (2019). Robust statistical methods in R using the WRS2 package. Behavior Research Methods. https://doi.org/10.3758/s13428-019-01246-w ] with a description of the functions currently available. (Here is the link for [direct access to the PDF](https://dornsife.usc.edu/assets/sites/239/docs/WRS2.pdf).)

Whenever a function of the CRAN version is used you can conveniently use the RStudio help utility to look into the details of this function.

## Summary

There are several options to load the required packages. Each has some advantages and disadvantages:

| Option | Advantage | Disadvantage |
|--------|-----------|--------------|
| `Rallfun-v??.txt` | All functions <br />Up to date, latest version <br/>Help infos available | Memory problem <br>Not included as R packages <br/>Own environment to prevent obstracting workspace <br />Help infos as comments baked into function |
| `WRS` (Github) | Many functions <br />R package | Not all functions<br/>Not obvious which are missing<br />No help pages |
| `WRS2` (CRAN) | R package<br />Help files | Only a small subset of functions |

To get a better overview which functions are used and to have full control of using it I will use a double strategy to implement these functions of the Wilcox book: 

1) I will use the file `Rallfun-v??.txt` to get all functions into the R memeory. But I will create a separate environment `allfun` to prevent memory problems and the clogging of the workspace. For using these special functions provided by the book, I will have to address this special environment by the `with` command.

2) I will note missing functions from `Rallfun-v??.txt` and looking for them in the other files and/or in the older version of the `Rallfun-v??.txt`-file. I will add these functions to the `allfun` environment.^[In a previous draft version of the first chapters I had displayed the source code of all functions and also my program code for administrative issues, e.g., the code how an external graphic is centered deisplayed. In this new version I have abstained from the additional complexity how to use the `bookdown` package and have concentrated on statistical issues.]


3) Finally at the end of each chapter I will remove all non-base packages and clean up the working space so that I can start each chapter from scratch without side effects.

```{r cleanup, results='hide', include=FALSE}
# xxx <- suppressWarnings(lapply(names(sessionInfo()$otherPkgs),
#                         function(pkgs)
#                           detach(
#                             paste0('package:', pkgs),
#                             character.only = T,
#                             unload = T,
#                             force = T
#                           )))
# 

rm(list = setdiff(ls(), freeze))
```

